Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2022-12-30 manuscript.tex
:pencil: fig intr
:pencil: intro str

Recent years have seen a striking increase in the number and variety of research studies that follow participants' everyday experiences and collect real-world psychological time series \citep[e.g.,][; also see \fgrref{fig:ScopusEsm}]{hamaker2017}\footnote{In psychology, extensive longitudinal data collection methods are often referred to as experience sampling method (ESM), ecological momentary assessment (EMA), or ambulatory assessment (AA) studies. While the terms come for different conceptual backgrounds, they share a focus on collecting data over an extended period of time capturing people's behaviors and experiences as they vary over time and in response to different situations and events. In this article, we will use the experience sampling (ESM) term as it has the strongest footing within the clustering literature.}. These extensive longitudinal data sets come with new forms of heterogeneity, where researchers have to consider differences across large numbers of participants, time points, and variables \citep[e.g.,][]{cattell1966, Monden2015, wardenaar2013}. Oftentimes, however, researchers are interested in precisely this complexity and wish to understand how people differ in their temporal developments across several variables \citep[i.e., understand the structure of between-subject differences of within-participant dynamics; e.g.,][]{ernst2021}. One relatively recent addition to the toolbox of ESM researchers has been time series clustering --- the idea of grouping participants based on the similarity of their time series \citep{ariens2020}.

@@

Time series clustering has a number of conceptual use cases with psychological data. Prime among them is the ability to reduce the time, variable, and person complexity by extracting structure between participants. These reduction and structuring qualities can be essential in detecting phenomena and extracting more abstract functional principles \citep[][]{eronen2021a}. These phenomena and principles can be meaningful differences between participants in different clusters, as well important patterns, trends, and relationships that participants share within a cluster \citep[e.g.,][]{schrodt2000}. Once distinct groups and patterns have been identified, researchers can examine the extent to which these within-group and between-group structures are associated with other variables of interest, such as personality traits, demographic characteristics, or other psychological constructs \citep[e.g.,][]{monden2022}. Once meaningful and robust structures and patterns are detected, time series clustering can, thus, be used to inform the development of robust theories as well as targeted interventions and therapies for individuals, for example, with mood disorders and other psychological conditions \citep[e.g.,][]{borsboom2021, eronen2020}.

@@

The main aim of time series clustering is to inductively group participants in such a way that the dynamics of some variable(s) have maximum similarity for individuals in the same group, and minimum similarity for individuals in different groups \citep[e.g.,][]{liao2005}. In practice, this means time series clustering critically depends on identifying similarities or dissimilarities in time series and then accurately distinguishing different groups (REF).

>>>>

Recent years have seen a striking increase in the number and variety of research studies that follow participants' everyday experiences and collect real-world psychological time series \citep[e.g.,][; also see \fgrref{fig:ScopusEsm}]{hamaker2017}\footnote{In psychology, extensive longitudinal data collection methods are often referred to as experience sampling method (ESM), ecological momentary assessment (EMA), or ambulatory assessment (AA) studies. While the terms come for different conceptual backgrounds, they share a focus on collecting data over an extended period of time to capture people's behaviors and experiences as they vary over time and in response to different situations and events. In this article, we will use the experience sampling (ESM) term as it has the strongest footing within the clustering literature.}. These extensive longitudinal data sets come with new forms of heterogeneity, where researchers have to consider differences across large numbers of participants, time points, and variables \citep[e.g.,][]{cattell1966, Monden2015, wardenaar2013}. Oftentimes, however, researchers are interested in precisely this complexity and wish to understand how people differ in their temporal developments across several variables \citep[e.g.,][]{ernst2021}. One relatively recent addition to the toolbox of ESM researchers has been time series clustering --- the idea of inductively grouping participants based on the similarity of their time series \citep{ariens2020}.

@@

Time series clustering has a number of conceptual use cases with psychological data. Prime among them is the ability to reduce the time, variable, and person complexity by extracting participant-level structures. These reduction and structuring qualities can be essential in detecting phenomena and extracting more abstract functional principles \citep[][]{eronen2021a}. These phenomena and principles can be meaningful differences that distinguish participants in different clusters, as well important patterns, trends, and relationships that participants share within a cluster \citep[e.g.,][]{schrodt2000}. Once distinct groups and patterns have been identified, researchers can examine the extent to which these within-group and between-group structures are associated with other variables of interest, such as personality traits, demographic characteristics, or other psychological constructs \citep[e.g.,][]{monden2022}. Once meaningful and robust structures and patterns are detected, time series clustering can, thus, be used to inform the development of robust theories as well as targeted interventions and therapies for individuals, for example, with mood disorders and other psychological conditions \citep[e.g.,][]{borsboom2021, eronen2020}.

@@

The main aim of time series clustering is to inductively group participants in such a way that the dynamics of some variable(s) have maximum similarity for individuals in the same group, and minimum similarity for individuals in different groups \citep[e.g.,][]{Aghabozorgi2015}. In practice, this means time series clustering critically depends on identifying similarities or dissimilarities in time series and then accurately distinguishing different groups \citep[e.g.,][]{Aghabozorgi2015}. Broadly speaking, time series (dis-)similarity can be approximated by either comparing the raw time series or by comparing some numerical representation of the time series (e.g., \citealp[]{liao2005}; see \fgrref{fig:tsClustTax} for an illustration). The (dis-)similarity measurements can then be clustered either on their own or as part of the representation estimation (e.g., \citealp[]{denteuling2021}; see `embedded' in \fgrref{fig:tsClustTax}).

\begin{figure}[!hbtp] %hbtp
  \caption{Time Series Clustering Taxonomy}
  \label{fig:tsClustTax}
  \centering\includegraphics[width=\textwidth]{figures/TS Cluster Flow/tsClustTax.pdf}
  \caption*{Note: \\
  The taxonomy only exemplifies some of the basic differences between a number of common time series clustering approaches. As such, the taxonomy and the notes are neither exhaustive nor complete in distinguishing different approaches. Additionally, terms and labels are used inconsistently across different types of literature and are chosen to avoid overlapping labels.}
\end{figure}
